---
title: "Data Processing Functions"
author: "Josh Pearman"
date: "2024-12-10"
output: html_document
---

Below is a collection of data processing and cleaning functions for data types I often work with, ranging from managing time variables, expanding survey-based network items stored in a concatenated list into a full column, and working with bibliometric scientific data. I'll return to this at some point in the future for documentation and refactoring, and maybe add some examples. I'll also have to rework the functions so that variables are set as parameters, since right now they're built into the functions themselves for my specific use-cases.


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Network analysis data

These functions are designed to standardize network items commonly administered on Qualtrics or other survey websites. For example, if you ask 5 people to rate their team members and the data is stored in a concatenated list (e.g., "5, 4, 2"), these functions will expand that to make it easier for extracting an adjacency matrix.

```{r network}
# Standardize network items into list format
process_column <- function(column) {
  sapply(column, function(cell) {
    if (grepl("^c\\(", cell)) {
      # If the cell starts with 'c(', it's likely a concatenated list
      numbers <- eval(parse(text = cell))
      paste(numbers, collapse = ", ")
    } else {
      # If not, return the original value
      cell
    }
  })
}
# Customizable function for using process_column function on network items by suffix
process_network_columns <- function(df) {
  # Get all column names
  all_cols <- names(df)
  
  # Find columns ending with "LIMIT_1"
  limit_1_cols <- all_cols[endsWith(all_cols, "LIMIT_1")]
  
  # Process each LIMIT_1 column
  for (col in limit_1_cols) {
    df[[col]] <- process_column(df[[col]])
  }
  
  return(df)
}

# Turn network items into matrix format
expand_column <- function(df, column_name, new_col_prefix, maxcol) {
  # Create a matrix to store the expanded column values
  expanded_values <- matrix(0, nrow = nrow(df), ncol = maxcol)
  maxcol_name <- maxcol-1
  # Loop through each row of your data frame and fill in the appropriate values in expanded_values
  for (i in 1:nrow(df)) {
    column_values <- strsplit(as.character(df[[column_name]][i]), ",")[[1]]
    for (j in 1:length(column_values)) {
      expanded_values[i, as.numeric(column_values[j])] <- 1
    }
  }
  
  # Add the expanded columns to your data frame with the given prefix
  new_colnames <- paste0(new_col_prefix, "_", 1:maxcol)
  df <- cbind(df, expanded_values)
  names(df)[(ncol(df)-maxcol_name):ncol(df)] <- new_colnames
  
  # Return the updated data frame
  return(df)
}


```

# Time functions

These are functions for various time-related purposes, such as splitting time data into different buckets, pivoting time data from long to wide for time-series analysis, and generating new columns backwards or forwards based on an input time (which helps if you're trying to create a standardized time dataframe across multiple experiments that occurred at different days and times)

```{r time}
# Function to extract times and create 10min, 20min, and 30min labels
extract_times <- function(input_time) {
  # Parse input_time into a POSIXct object
  input_time <- strptime(input_time, format = "%H:%M")
  
  # Calculate start time, 10 min, 20 min, and 30 min after start
  start_time <- input_time - minutes(45)
  time_10min <- start_time + minutes(10)
  time_20min <- start_time + minutes(20)
  time_30min <- start_time + minutes(30)
  
  # Create a data frame with SurveyTime and Time columns
  result <- data.frame(
    SurveyTime = c(time_10min, time_20min, time_30min),
    Time = c(10, 20, 30)
  )
  
  # Format SurveyTime as HH:MM
  result$SurveyTime <- format(result$SurveyTime, format = "%H:%M")
  
  return(result)
}

# Function to compare times and update Time column
update_time_column <- function(df, survey_times, range_var) {
  # Convert hm_time to POSIXct format
  df$hm_time <- strptime(df$hm_time, format = "%H:%M")
  survey_times$SurveyTime <- strptime(survey_times$SurveyTime, format = "%H:%M")
  
  # Initialize an empty vector to store updated Time values
  updated_times <- numeric(nrow(df))
  
  # Iterate through each row in the dataframe
  for (i in 1:nrow(df)) {
    # Get the HM_time from the dataframe
    hm_time <- df$hm_time[i]
    
    # Calculate time differences between hm_time and SurveyTime values
    time_diffs <- as.numeric(difftime(hm_time, survey_times$SurveyTime, units = "mins"))
    
    # Find the index of the closest SurveyTime within 10 minutes
    closest_index <- which(time_diffs < range_var)
    
    # If a matching SurveyTime within 10 minutes is found, update Time column
    if (length(closest_index) > 0) {
      updated_times[i] <- survey_times$Time[closest_index[1]]
    } else {
      updated_times[i] <- NA  # Set Time to NA if no match is found within 10 minutes
    }
  }
  
  # Update the df$Time column with the matched Time values
  df$Time <- updated_times
  
  return(df)
}

# Function version of code to pivot data using time

time_wide <- function(df) {
  df <- df %>%
    # Create time variable
    mutate(time = format(as.POSIXct(creationDateTime), format = "%H:%M:%S")) %>%
    # Remove deletion coordinate behaviors
    filter(type != "DELETEMAPCELL" & type != "DELETESAVEDCALCULATION") %>% 
    # Sort by time variable so indexing is correct
    arrange(time)
  # Create index var
  df$index <- 1:nrow(df)
  df <- df %>% 
    # Sort downwards by index so that last-selected coordinates are first
    arrange(desc(index)) %>% 
    # Only retain distinct coordinates by sender
    # Specifically retains last-selected coordinates
    distinct(sender, time, .keep_all = TRUE) %>%
    # Remove index variable
    select(-c(index)) %>% 
    # Rearrange chronologically
    arrange(time)
  
  # Pivot wider by time variable
  wide_df <- df %>%
    pivot_wider(
      id_cols = c("campaign", "mission", "session", "sender"),
      names_from = time,
      values_from = coordinate
    )
  return(wide_df)
}

## Function to take input start time and generate an empty dataframe with one second
## per column for 45 minutes, or 2700 columns.

time_gen_forward <- function(start_time) {
  convert_time <- strptime(start_time, format = "%H:%M:%S")
  time_vector <- seq(convert_time, by = "1 sec", length.out = 2700)
  
  time_vector <- format(time_vector,"%H:%M:%S")
  
  df <- data.frame(matrix(ncol = length(time_vector) + 4, nrow = 0))
  colnames(df) <- c("campaign", "mission", "session", "sender", as.character(time_vector))
  
  return(df)
}

## Function to take input end time and generate an empty dataframe with one second
## per column for 45 minutes, or 2700 columns.

time_gen_backward <- function(end_time) {
  convert_time <- strptime(end_time, format = "%H:%M:%S") - as.difftime(2700, unit = "secs")
  time_vector <- seq(convert_time, by = "1 sec", length.out = 2700)
  
  time_vector <- format(time_vector,"%H:%M:%S")
  
  df <- data.frame(matrix(ncol = length(time_vector) + 4, nrow = 0))
  colnames(df) <- c("campaign", "mission", "session", "sender", as.character(time_vector))
  
  return(df)
}

## Function to merge time name column with the wide format dataframe

merge_times <- function(df_time, wide_df) {
  df <- data.frame(matrix(ncol = ncol(df_time), nrow = nrow(wide_df)))
  colnames(df) <- colnames(df_time)
  
  for (col in colnames(wide_df)) {
    if (col %in% colnames(df)) {
      df[[col]] <- wide_df[[col]]
    } else {
      df[[col]] <- NA
    }
  }
  
  df[df == "NULL"] <- NA
  
  # This keeps only up to the 45th minute (or 2700th second) of a session.
  # Specifically, use the MTS signoff time and then generate 2700 seconds backwards from there
  # to determine start time
  df <- df[,c(1:2704)] 
  
  return(df)
}


```

# Text formatting code

Here are some functions that I use to extract variable names and item content from a standard Qualtrics survey (in case you want to analyze items directly), a function for fuzzy matching between two string columns (useful when trying to compare all the different names one university might go by), and a function for extracting the first and last authors of a paper that is flexible across team size.

```{r text-cleaning}
# Function to clean the text from Qualtrics items
clean_text <- function(text) {
  # Remove everything up to and including the dash and space after it
  cleaned <- sub("^.*- ", "", text)
  return(cleaned)
}

# Function to find and replace item labels with texts
replace_with_text <- function(keys, texts) {
  match_col <- which(names(texts) == keys)
  if (length(match_col) > 0) {
    return(texts[1, match_col])  # Assuming the text is in the first row
  } else {
    return(keys)  # Return original value if no match found
  }
}

# Fuzzy string matching functions
# Function for fuzzy string matching of university names
standardize_name <- function(x) {
  str_to_lower(x) %>%
    # Remove common separators
    str_replace_all("[,\\-\\_]", " ") %>%
    # Remove multiple spaces
    str_replace_all("\\s+", " ") %>%
    str_trim()
}

# Function to identify first and last authors in a bibliometric dataset
identify_first_last <- function(df) {
  # Split coauthor names by semicolon (;)
  coauthors <- strsplit(df$coauthor_names, ";")
  
  # Iterate over each row
  for (i in 1:nrow(df)) {
    # Split the coauthors for the current row by comma (,)
    names <- strsplit(coauthors[[i]], ",")
    
    # Trim leading/trailing whitespace for each name
    names <- lapply(names, trimws)
    
    # Check if the first name matches the firstname column
    if (names[[1]][1] == df$last_name[i] & names[[1]][2] == df$first_name[i]) {
      df$authorship_order[i] <- "first"
    }
    
    # Check if the last name matches the lastname column
    else if (names[[length(names)]][1] == df$last_name[i] & names[[length(names)]][2] == df$first_name[i]) {
      df$authorship_order[i] <- "last"
    }
    
    else{
      df$authorship_order[i] <- "middle"
    }
  }
  return(df)
}
```

